\documentclass[main.tex]{subfiles}
\begin{document}
	\begin{abstract}
							
		GRIN is short for Graph Reduction Intermediate Notation~\cite{boquist-phd}, a modern back-end for lazy functional languages. Most of the currently available compilers for such languages share a common flaw: they can only optimize programs on a per-module basis. The GRIN framework allows for interprocedural whole program analysis, enabling optimizing code transformations across functions and modules as well.
		
		Some implementations of GRIN already exist, but most of them were developed only for experimentation purposes. Thus, they either compromise on low level efficiency, or contain ad hoc modifications compared to the original specification.
		
		Our goal is to provide a full-fledged implementation of GRIN by combining the currently available best technologies like LLVM, and measure the framework's effectiveness compared to some of the most well-known functional language compilers such as the Glasgow Haskell Compiler and the Idris compiler. We also present some improvements to the already existing components of the framework. Some of these improvements include a typed representation for the intermediate language and a new optimization, the interprocedural dead data elimination.
		
		\vfill
		
				
		\hspace{3cm}
		
		\begin{center}
			The project has been supported by the European Union, \\
			co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).
		\end{center}
		
	\end{abstract}
\end{document}